---
title: "Understanding Farm Planting Dates in Zambia"
author: "Bryce Stouffer & Ryan Kopper"
date: "May 2, 2019"
output: html_document
---

## **Do farming communities vary their planting dates to adapt to rainfall uncertainty?**

### 1. Summary 
Do farming communities in areas with higher rainfall variability show higher variability between farmers in terms of when crops are planted? If so, this could portray a community-level adaptive response for dealing with climatic uncertainty.

This analysis will utilize remotely sensed datasets: NDVI time series, gridded rainfall datasets, and a cropland mask. The NDVI time series are processed into values that identify the date of green-up over ~18 years (our planting date proxy). The rainfall data will be used to calculate the coefficient of variation (CV) of rainfall in the first 2 months of the growing season, and the long-term mean rainfall. Comparing croplands from areas of similar mean rainfall, we will see if there is a relationship between CV of rainfall and CV of green-up date.

### Primary Objectives

 1. Obtain data (green-up and rainfall)
 2. Use rainfall data to: 
    + Calculate the coefficient of variation (CV) of weekly rainfall across the 2000-2018 growing season.
    + Calculate the long-term mean rainfall.
    + Divide Zambia into zones based on long-term mean rainfall.
 3. Use green-up data to:
    + Calculate the pixel-wise coefficient of variation (CV) of green-up in for the ~18 year time span (2000-2018) cropping seasons.
    + Calculate the pixel-wise mean of green-up for the ~18 year time span (2000-2018) cropping seasons. Then aggregate using CV function.
4. Determine if there is a relationship between CV of rainfall during the growing season and the CV of green-up date (pixel-wise or aggregate).

### Data and Libraries

Data for this project can be downloaded here:
https://drive.google.com/drive/folders/1OZIRP-M8agu4Gyq_HFAXqrwzyNMxffaD 

```{r}
library(geospaar)
library(googledrive)
library(data.table)
library(parallel)
library(foreach)
library(doParallel)
library(raster)
#Mac users
#library(doMC)

#reading in the planting dates data (green-up)
PD <- dir(path ="C:/Users/bstouffer/Desktop/Geospaar/plantingdatecv/data/Estimated_Planting_Date_2000-2018/", pattern = ".tif", full.names = TRUE)

#creating planting date raster
PD_ras <- lapply(PD, raster)
PD_stack <- brick(PD_ras)

#reading in cropmask data
cropmask <- raster(file.path("C:/Users/bstouffer/Desktop/Geospaar/plantingdatecv/data/CropZmb1V1_2015_16_fullagree.tif"))

#testing rainfall data
chrpsf <- list.files("C:/Users/bstouffer/Desktop/Geospaar/plantingdatecv/data/chirps", full.names = TRUE)
chrps <- lapply(chrpsf, brick)
```

### Processing the rainfall data

The rainfall data was aqcuired from Google Earth Engine. That code can be found
here: https://code.earthengine.google.com/4d8ad06c67e84615e61e4a9d539c1432. 
The rainfall data from Google Earth Engine is processed into weekly data from 
daily. Depending on your operating system, the libraries needed may vary. 
The code below is for Windows.

```{r}
#subsetting the rainfall data to the growing season
yrs <- 2000:2018
wks <- lapply(1:length(chrps), function(x) {  # x <- 1
  drng <- as.Date(paste0(c(yrs[x], yrs[x] + 1), "-", c("10-01", "01-15")))
  dts <- seq.Date(drng[1], drng[2], by = "day")
  if(length(dts) != nlayers(chrps[[x]])) stop("Mismatch", call. = FALSE)
  DT <- data.table("ind" = 0, "dts" = dts, "yrs" = year(dts), 
                   "wk" = isoweek(dts))
  DT[wk %in% c(1:2, 40:52), ind := 1]
})

#pulling out full weeks of rainfall data
detectCores()
registerDoParallel(8)

wkly_rfl <- foreach(i = 1:length(chrps)) %do% {
  wkly_rf <- lapply(c(40:52, 1:2), function(y) {  # i <- 10; y <- 44
    ind <- which(wks[[i]]$ind == 1 & wks[[i]]$wk == y)
    # wks[[i]][ind, ]
    r <- raster::calc(chrps[[i]][[ind]], sum)
  })
  s <- stack(wkly_rf)
}
```

### The CV for green-up calculations

Green-up dates are first masked to cropland in Zambia. Then they are calculated
and aggregated. 

```{r}
#projecting the cropland mask
cropmask_proj_green <- projectRaster(from = cropmask, to = PD_stack)

#reclassing mask based on probabilities of cropland
#cropland mask is based on another analysis where NDVI > .75 is considered cropland
cmask_green <- reclassify(cropmask_proj_green, c(.75, 1, 1, 0, .7499, 0))

#masking plantingdates to zambia
pdmasked <- mask(x = PD_stack, mask = cmask_green, maskvalue = 0)
pdmasked2 <- mask(x = pdmasked, mask = cmask_green)

#calculating the pixel-wise green-up CV
greencvnoagg <- calc(pdmasked2, cv) 

#aggregate to same spatial resolution of rainfall
greencv <- aggregate(x = greencvnoagg, fact = 2)
plot(greencv, main = "Green-up CV")

#greenup CV pixel wise mean, then aggregate with CV function
greenmean <- calc(pdmasked2, mean)
greenmeancv <- aggregate(x = greenmean, fact = 2, method = "cv")
plot(greenmeancv, main = "Green-up Mean Aggregated CV")
```

### Rainfall masked to cropland

```{r}
#masking rainfall to zambia's cropland
cropmask_proj <- resample(x = cropmask_proj_green, y = greencv)
cmask <- reclassify(cropmask_proj, c(.75, 1, 1, 0, .7499, 0))

wkly_rfl <- lapply(1:length(wkly_rfl), function(z) {
  resample(x = wkly_rfl[[z]], y = greencv)
})

rainfallmasked <- lapply(1:length(wkly_rfl), function(y) {
    mask(x = wkly_rfl[[y]], mask = cmask, maskvalue = 0)
})
rainfallmasked2 <- lapply(1:length(wkly_rfl), function(y) {
    mask(x = rainfallmasked[[y]], mask = cmask)
})

plot(rainfallmasked2[[1]][[1]], main = "Week 1 Rainfall Year 2000")
```

### Calculating longterm mean of annual rainfall

```{r}
#mean annual rainfall with mask
annual_rfl_masked <- calc(stack(lapply(rainfallmasked2, function(x) calc(x, sum))), mean)
plot(annual_rfl_masked)
```

### Calculating the mean weekly CV of rainfall across all growing seasons

```{r}
rf_weekly_cv_masked <- calc(stack(lapply(rainfallmasked2, function(x) {
  calc(x, cv)
})), mean)

plot(rf_weekly_cv_masked, main = "Mean Weekly CV Across All Growing Seasons")
```

### Examining the relationship between green-up and rainfall CV across all of Zambia

```{r}
#Linear model test, green-up CV (pixel-wise) & weekly CV average across all growing seasons
#Using cropland probability as weights
names(greencv) <- "greencv"

pts <- rasterToPoints(greencv, sp = TRUE) %>% st_as_sf
pts <- pts %>% mutate(rcv = raster::extract(x = rf_weekly_cv_masked, y = .))
pts <- pts %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat <- bind_cols(pts %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts) %>% as_tibble) %>% drop_na

greenlm <- lm(greencv ~ rcv, data = pts_dat, weights = pts_dat$wts)
summary(greenlm)
plot(greencv ~ rcv, data = pts_dat, ylab = "green-up cv", xlab = "rainfall cv")
```

### Dividing Zambia into zones based on longterm mean annual rainfall

The scale ranges from 1 (low) to 6 (high).

```{r}
qtiles <- quantile(annual_rfl_masked, probs = seq(0, 1, 1 / 6))
raintotcut <- cut(x = annual_rfl_masked, breaks = qtiles, include.lowest = TRUE)
plot(raintotcut)
```

### Examining the relationship between green-up CV in each rainfall zone

Cropland probability are used as weights for all models.

#### Zone 1

```{r}
#masking greenup and weekly rainfall based on longterm avg
noone <- subs(raintotcut, data.frame(from=c(2, 3, 4, 5, 6), 1))
greencvmask1 <- mask(x = greencv, mask = noone, maskvalue = 1)
plot(greencvmask1)
rf_weekly_cv_masked1 <- mask(x = rf_weekly_cv_masked, mask = noone, maskvalue = 1)
plot(rf_weekly_cv_masked1)

#lm for class 1 (low) longterm rainfall crop areas (pixel-wise green-up)
#using cropland probability as weights
names(greencvmask1) <- "greencvmask1"
pts1 <- rasterToPoints(greencvmask1, sp = TRUE) %>% st_as_sf
pts1 <- pts1 %>% mutate(rcv1 = raster::extract(x = rf_weekly_cv_masked1, y = .)) 
pts1 <- pts1 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat1 <- bind_cols(pts1 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts1) %>% as_tibble) %>% drop_na

greenlm1 <- lm(greencvmask1 ~ rcv1, data = pts_dat1, weights = pts_dat1$wts)
summary(greenlm1)
plot(greencvmask1 ~ rcv1, data = pts_dat1)
plot(greencvmask1 ~ rcv1, data = pts_dat1, ylab = "green-up cv", xlab = "rainfall cv")
```

#### Using green-up mean (aggregated CV)

```{r}
#masking greenup and weekly rainfall based on longterm avg
noone <- subs(raintotcut, data.frame(from=c(2, 3, 4, 5, 6), 1))
greenmeancvmask1 <- mask(x = greenmeancv, mask = noone, maskvalue = 1)
plot(greenmeancvmask1)
rf_weekly_cv_masked1 <- mask(x = rf_weekly_cv_masked, mask = noone, maskvalue = 1)
plot(rf_weekly_cv_masked1)

#class 1
names(greenmeancvmask1) <- "greenmeancvmask1"
pts1 <- rasterToPoints(greenmeancvmask1, sp = TRUE) %>% st_as_sf
pts1 <- pts1 %>% mutate(rcv1 = raster::extract(x = rf_weekly_cv_masked1, y = .)) 
pts1 <- pts1 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat1 <- bind_cols(pts1 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts1) %>% as_tibble) %>% drop_na

#lm for class 1 (low) longterm rainfall crop areas
greenmeanlm1 <- lm(greenmeancvmask1 ~ rcv1, data = pts_dat1, weights = pts_dat1$wts)
summary(greenmeanlm1)
plot(greenmeancvmask1 ~ rcv1, data = pts_dat1, ylab = "green-up cv", xlab = "rainfall cv")
```


#### Zone 2

```{r}
#class 2
notwo <- subs(raintotcut, data.frame(from=c(1, 3, 4, 5, 6), 1))
greencvmask2 <- mask(x = greencv, mask = notwo, maskvalue = 1)
plot(greencvmask2)
rf_weekly_cv_masked2 <- mask(x = rf_weekly_cv_masked, mask = notwo, maskvalue = 1)
plot(rf_weekly_cv_masked2)

#lm for class 2 longterm rainfall crop areas (pixel-wise green-up)
names(greencvmask2) <- "greencvmask2"
pts2 <- rasterToPoints(greencvmask2, sp = TRUE) %>% st_as_sf
pts2 <- pts2 %>% mutate(rcv2 = raster::extract(x = rf_weekly_cv_masked2, y = .)) 
pts2 <- pts2 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat2 <- bind_cols(pts2 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts2) %>% as_tibble) %>% drop_na

greenlm2 <- lm(greencvmask2 ~ rcv2, data = pts_dat2, weights = pts_dat2$wts)
summary(greenlm2)
plot(greencvmask2 ~ rcv2, data = pts_dat2, ylab = "green-up cv", xlab = "rainfall cv")
```

#### Using green-up mean (aggregated CV)

```{r}
#class 2
notwo <- subs(raintotcut, data.frame(from=c(1, 3, 4, 5 ,6), 1))
greenmeancvmask2 <- mask(x = greenmeancv, mask = notwo, maskvalue = 1)
plot(greenmeancvmask2)
rf_weekly_cv_masked2 <- mask(x = rf_weekly_cv_masked, mask = notwo, maskvalue = 1)
plot(rf_weekly_cv_masked2)

#lm for class 2  longterm rainfall crop areas
names(greenmeancvmask2) <- "greenmeancvmask2"
pts2 <- rasterToPoints(greenmeancvmask2, sp = TRUE) %>% st_as_sf
pts2 <- pts2 %>% mutate(rcv2 = raster::extract(x = rf_weekly_cv_masked2, y = .))
pts2 <- pts2 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat2 <- bind_cols(pts2 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts2) %>% as_tibble) %>% drop_na

greenmeanlm2 <- lm(greenmeancvmask2 ~ rcv2, data = pts_dat2, weights = pts_dat2$wts)
summary(greenmeanlm2)
plot(greenmeancvmask2 ~ rcv2, data = pts_dat2, ylab = "green-up cv", xlab = "rainfall cv")
```

#### Zone 3

```{r}
#class 3
nothree <- subs(raintotcut, data.frame(from=c(1, 2, 4, 5, 6), 1))
greencvmask3 <- mask(x = greencv, mask = nothree, maskvalue = 1)
plot(greencvmask3)
rf_weekly_cv_masked3 <- mask(x = rf_weekly_cv_masked, mask = nothree, maskvalue = 1)
plot(rf_weekly_cv_masked3)

#lm for class 3 longterm rainfall crop areas (pixel-wise green-up)
names(greencvmask3) <- "greencvmask3"
pts3 <- rasterToPoints(greencvmask3, sp = TRUE) %>% st_as_sf
pts3 <- pts3 %>% mutate(rcv3 = raster::extract(x = rf_weekly_cv_masked3, y = .)) 
pts3 <- pts3 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat3 <- bind_cols(pts3 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts3) %>% as_tibble) %>% drop_na

greenlm3 <- lm(greencvmask3 ~ rcv3, data = pts_dat3, weights = pts_dat3$wts)
summary(greenlm3)
plot(greencvmask3 ~ rcv3, data = pts_dat3, ylab = "green-up cv", xlab = "rainfall cv")

```

#### Using green-up mean (aggregated CV)

```{r}
#class 3
nothree <- subs(raintotcut, data.frame(from=c(1, 2, 4, 5, 6), 1))
greenmeancvmask3 <- mask(x = greenmeancv, mask = nothree, maskvalue = 1)
plot(greenmeancvmask3)
rf_weekly_cv_masked3 <- mask(x = rf_weekly_cv_masked, mask = nothree, maskvalue = 1)
plot(rf_weekly_cv_masked3)

#lm for class 3  longterm rainfall crop areas
names(greenmeancvmask3) <- "greenmeancvmask3"
pts3 <- rasterToPoints(greenmeancvmask3, sp = TRUE) %>% st_as_sf
pts3 <- pts3 %>% mutate(rcv3 = raster::extract(x = rf_weekly_cv_masked3, y = .))
pts3 <- pts3 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat3 <- bind_cols(pts3 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts3) %>% as_tibble) %>% drop_na

greenmeanlm3 <- lm(greenmeancvmask3 ~ rcv3, data = pts_dat3, weights = pts_dat3$wts)
summary(greenmeanlm3)
plot(greenmeancvmask3 ~ rcv3, data = pts_dat3, ylab = "green-up cv", xlab = "rainfall cv")
```

#### Zone 4

```{r}
#class 4
nofour <- subs(raintotcut, data.frame(from=c(1, 2, 3, 5, 6), 1))
greencvmask4 <- mask(x = greencv, mask = nofour, maskvalue = 1)
plot(greencvmask4)
rf_weekly_cv_masked4 <- mask(x = rf_weekly_cv_masked, mask = nofour, maskvalue = 1)
plot(rf_weekly_cv_masked4)

#lm for class 4  longterm rainfall crop areas (pixel-wise green-up)
names(greencvmask4) <- "greencvmask4"
pts4 <- rasterToPoints(greencvmask4, sp = TRUE) %>% st_as_sf
pts4 <- pts4 %>% mutate(rcv4 = raster::extract(x = rf_weekly_cv_masked4, y = .))
pts4 <- pts4 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat4 <- bind_cols(pts4 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts4) %>% as_tibble) %>% drop_na

greenlm4 <- lm(greencvmask4 ~ rcv4, data = pts_dat4, weights = pts_dat4$wts)
summary(greenlm4)
plot(greencvmask4 ~ rcv4, data = pts_dat4, ylab = "green-up cv", xlab = "rainfall cv")
```

#### Using green-up mean (aggregated CV)

```{r}
#class 4
nofour <- subs(raintotcut, data.frame(from=c(1, 2, 3, 5, 6), 1))
greenmeancvmask4 <- mask(x = greenmeancv, mask = nofour, maskvalue = 1)
plot(greenmeancvmask4)
rf_weekly_cv_masked4 <- mask(x = rf_weekly_cv_masked, mask = nofour, maskvalue = 1)
plot(rf_weekly_cv_masked4)

#lm for class 4  longterm rainfall crop areas 
names(greenmeancvmask4) <- "greenmeancvmask4"
pts4 <- rasterToPoints(greenmeancvmask4, sp = TRUE) %>% st_as_sf
pts4 <- pts4 %>% mutate(rcv4 = raster::extract(x = rf_weekly_cv_masked4, y = .))
pts4 <- pts4 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat4 <- bind_cols(pts4 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts4) %>% as_tibble) %>% drop_na

greenmeanlm4 <- lm(greenmeancvmask4 ~ rcv4, data = pts_dat4, weights = pts_dat4$wts)
summary(greenmeanlm4)
plot(greenmeancvmask4 ~ rcv4, data = pts_dat4, ylab = "green-up cv", xlab = "rainfall cv")
```

#### Zone 5 

```{r}
#class 5
nofive <- subs(raintotcut, data.frame(from=c(1, 2, 3, 4, 6), 1))
greencvmask5 <- mask(x = greencv, mask = nofive, maskvalue = 1)
plot(greencvmask5)
rf_weekly_cv_masked5 <- mask(x = rf_weekly_cv_masked, mask = nofive, maskvalue = 1)
plot(rf_weekly_cv_masked5)

#lm for class 5 longterm rainfall crop areas (pixel-wise green-up)
names(greencvmask5) <- "greencvmask5"
pts5 <- rasterToPoints(greencvmask5, sp = TRUE) %>% st_as_sf
pts5 <- pts5 %>% mutate(rcv5 = raster::extract(x = rf_weekly_cv_masked5, y = .))
pts5 <- pts5 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat5 <- bind_cols(pts5 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts5) %>% as_tibble) %>% drop_na

greenlm5 <- lm(greencvmask5 ~ rcv5, data = pts_dat5, weights = pts_dat5$wts)
summary(greenlm5)
plot(greencvmask5 ~ rcv5, data = pts_dat5, ylab = "green-up cv", xlab = "rainfall cv")
```

#### Using green-up mean (aggregated CV)

```{r}
#class 5
nofive <- subs(raintotcut, data.frame(from=c(1, 2, 3, 4, 6), 1))
greenmeancvmask5 <- mask(x = greenmeancv, mask = nofive, maskvalue = 1)
plot(greenmeancvmask5)
rf_weekly_cv_masked5 <- mask(x = rf_weekly_cv_masked, mask = nofive, maskvalue = 1)
plot(rf_weekly_cv_masked5)

#lm for class 5 longterm rainfall crop areas
names(greenmeancvmask5) <- "greenmeancvmask5"
pts5 <- rasterToPoints(greenmeancvmask5, sp = TRUE) %>% st_as_sf
pts5 <- pts5 %>% mutate(rcv5 = raster::extract(x = rf_weekly_cv_masked5, y = .))
pts5 <- pts5 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat5 <- bind_cols(pts5 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts5) %>% as_tibble) %>% drop_na

greenmeanlm5 <- lm(greenmeancvmask5 ~ rcv5, data = pts_dat5, weights = pts_dat5$wts)
summary(greenmeanlm5)
plot(greenmeancvmask5 ~ rcv5, data = pts_dat5, ylab = "green-up mean cv", xlab = "rainfall cv")
```

#### Zone 6

```{r}
#class 6
nosix <- subs(raintotcut, data.frame(from=c(1, 2, 3, 4, 5), 1))
greencvmask6 <- mask(x = greencv, mask = nosix, maskvalue = 1)
plot(greencvmask6)
rf_weekly_cv_masked6 <- mask(x = rf_weekly_cv_masked, mask = nosix, maskvalue = 1)
plot(rf_weekly_cv_masked6)

#lm for class 6 (high) longterm rainfall crop areas (pixel-wise green-up)
names(greencvmask6) <- "greencvmask6"
pts6 <- rasterToPoints(greencvmask6, sp = TRUE) %>% st_as_sf
pts6 <- pts6 %>% mutate(rcv6 = raster::extract(x = rf_weekly_cv_masked6, y = .))
pts6 <- pts6 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat6 <- bind_cols(pts6 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts6) %>% as_tibble) %>% drop_na

greenlm6 <- lm(greencvmask6 ~ rcv6, data = pts_dat6, weights = pts_dat6$wts)
summary(greenlm6)
plot(greencvmask6 ~ rcv6, data = pts_dat6, ylab = "green-up cv", xlab = "rainfall cv")
```

#### Using green-up mean (aggregated CV)

```{r}
#class 6
nosix <- subs(raintotcut, data.frame(from=c(1, 2, 3, 4, 5), 1))
greenmeancvmask6 <- mask(x = greenmeancv, mask = nosix, maskvalue = 1)
plot(greenmeancvmask6)
rf_weekly_cv_masked6 <- mask(x = rf_weekly_cv_masked, mask = nosix, maskvalue = 1)
plot(rf_weekly_cv_masked6)

#lm for class 6 (high) longterm rainfall crop areas
names(greenmeancvmask6) <- "greenmeancvmask6"
pts6 <- rasterToPoints(greenmeancvmask6, sp = TRUE) %>% st_as_sf
pts6 <- pts6 %>% mutate(rcv6 = raster::extract(x = rf_weekly_cv_masked6, y = .)) 
pts6 <- pts6 %>% mutate(wts = raster::extract(x = cropmask_proj, y = .))
pts_dat6 <- bind_cols(pts6 %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts6) %>% as_tibble) %>% drop_na

greenmeanlm6 <- lm(greenmeancvmask6 ~ rcv6, data = pts_dat6, weights = pts_dat6$wts)
summary(greenmeanlm6)
plot(greenmeancvmask6 ~ rcv6, data = pts_dat6, ylab = "green-up mean cv", xlab = "rainfall cv")
```


### Conclusion

There does not appear to be a relationship between the CV of green-up (planting
date) and the CV of rainfall. The two differing methods of green-up CV calculations did not lend differing results. However, our analysis was carried out at 500 meter resolution. Different scales of analysis could yield different results, especially for smaller community farms. 
